Towards Verifying the Bitcoin-S Library

### 2019-06-26 paper/paper.tex
start working on transformations

These are the code transformations needed to Verify the Addition-with-Zero Property.

When we run Stainless on this code (without any properties to prove), it throws the following errors:

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:6:3: Stainless doesn't support abstract type members
            type A
            ^^^^^^
[ Error  ] currency/CurrencyUnits.scala:26:33: Only literal arguments are allowed for BigInt.
            def toBigInt: BigInt = BigInt(toLong)
                                          ^^^^^^
[ Error  ] currency/CurrencyUnits.scala:33:1: Objects cannot extend classes or implement traits, use a case object instead
          object Satoshis extends BaseNumbers[Satoshis] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[ Error  ] number/NumberType.scala:65:1: Objects cannot extend classes or implement traits, use a case object instead
          object Int64 extends BaseNumbers[Int64] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[  Info  ] Shutting down executor service.
\end{lstlisting}

So let's see how we can fix those errors.

\subsection{An Inheriting Object}

Stainless output:
\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:33:1: Objects cannot extend classes or implement traits, use a case object instead
          object Satoshis extends BaseNumbers[Satoshis] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[ Error  ] number/NumberType.scala:65:1: Objects cannot extend classes or implement traits, use a case object instead
          object Int64 extends BaseNumbers[Int64] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
\end{lstlisting}

Code before:
\begin{lstlisting}[style=scala]
object Int64 extends BaseNumbers[Int64] { ... }
object Satoshis extends BaseNumbers[Satoshis] { ... }
\end{lstlisting}

Code after:
\begin{lstlisting}[style=scala]
case object Int64 extends BaseNumbers[Int64] { ... }
case object Satoshis extends BaseNumbers[Satoshis] { ... }
\end{lstlisting}

Here, we can just change the objects from object to case object.
Stainless recommendation is to use objects for modules and case objects as algebraic data types.

\todo{Kai this does not change the semantic}

This is due to the internal design of Scala.
It's possible to reason about case object but not about object.
This needs a fundamental knowledge of Scala and some functional paradigms that should not be part of this thesis.
The \href{https://github.com/epfl-lara/stainless/issues/520}{issue number 520} on Stainless GitHub gives some thoughts, if you want to know more.

\subsection{Abstract Type Member}

Stainless output:
\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:6:3: Stainless doesn't support abstract type members
            type A
            ^^^^^^
\end{lstlisting}

@@

\todo{kai this does not change the semantic}

@@

Satoshis extends CurrencyUnit with type Int64, so it would be of type CurrencyUnit[Int64].
That's too specific, because the return type of the addition is then CurrencyUnit[Int64] not CurrencyUnit[A].
Maybe the Bitcoin-S developers should reimplement this part and not use Satoshis directly.

Since there is no easy way to fix it and the code should stay as much as possible the original, we just remove the abstract type and set it to Int64.
This limits the verification a bit, but as we only want to verify the addition in satoshis, that's OK.

>>>>

Here we see in detail how to transform the bitcoin-s code into the
Scala fragment supported by Stainless. All subsections start with the
code on which Stainless is called, followed by the Stainless error
message(s) and finally a description of the changes we make to the
code.

We claim that all transformations are equivalent in the sense that if
the addition-with-zero property holds for the transformed code, then
it also holds for the code before the transformation.

\subsection{Inheriting Objects}

\lstinputlisting[style=scala]{../code/addition/src/main/scala/addition/reduced/currency/CurrencyUnits.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/addition/reduced/number/NumberType.scala}

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:33:1: Objects cannot extend classes or implement traits, use a case object instead
          object Satoshis extends BaseNumbers[Satoshis] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[ Error  ] number/NumberType.scala:65:1: Objects cannot extend classes or implement traits, use a case object instead
          object Int64 extends BaseNumbers[Int64] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
\end{lstlisting}

Here, we can just turn the objects into case objects by literally just
changing the word \texttt{object} into \texttt{case object} on lines
33 and 65 in the two respective files.

That transformation is equivalent. Case objects have some additional
properties (in particular, being serializable) and they inherit from
\texttt{Product} instead of \texttt{AnyRef}, but none of our code
depends on any of that.

\subsection{Abstract Type Members}

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:6:3: Stainless doesn't support abstract type members
            type A
            ^^^^^^
\end{lstlisting}

@@

@@

Satoshis extends CurrencyUnit with type Int64, so it would be of type CurrencyUnit[Int64].
That's too specific, because the return type of the addition is then CurrencyUnit[Int64] not CurrencyUnit[A].

Maybe the Bitcoin-S developers should reimplement this part and not use Satoshis directly.
\todo{ramon: explain. what would be the correct code?}

Since we only want to verify the addition of satoshis, and the Satoshis class overrides A with Int64 anyway,  we just remove the abstract type and set it to Int64.
